Picky with peakpicking: assessing chromatographic peak quality with simple

### 2023-06-07 manuscript/manuscript.Rmd
Updated version of Figure 5 with histograms and boxplots

A different story emerged, however, when the bad MFs were removed from this analysis. Good features were most commonly found to have their highest concentrations at the DCM or the surface, rather than being fixed with respect to depth. Of the 182 good MFs, less than a fifth had no trend with depth (44/249) and a majority had unequivocally lowest values in the 175 meter samples (those with 15m/DCM > 175m, 145 features). The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad [Figure 7B, currently in Slack?].

Additionally, a large number of features manually identified as bad nonetheless had significant differences with depth. This was surprising because we had assumed that bad MFs corresponded to instrument noise, which we did not expect to have any biological trend. Further investigation of a few randomly selected bad features with a biological difference revealed the reason behind this: most of those investigated were actually tails of other MFs. Integrating just the tail of a peak retains the biological signal of the full peak while still looking visually like instrument noise, thereby introducing pseudoreplication in the feature space.

@@

```{r Fig5 clockplot}
resp_class_levels <- c("All equal", 
    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
    "Hard to tell")
clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
  mutate(resp_class=factor(resp_class, levels=resp_class_levels))

@@

clockplot_base_gp <- ggplot() +
  annotate("segment", x=c(0,0,0), y=c(0,0,0), 
           xend=axis_xends, yend=axis_yends,
           arrow=arrow(length=unit(0.02, "npc"), type = "closed")) +
  annotate("segment", x=c(0,0,0), y=c(0,0,0),
           xend=-axis_xends, yend=-axis_yends,
           lty = 2) +
  annotate("label", x=axis_xends, y=axis_yends+axis_y_dodge,
           label=c("15m axis", "DCM axis", "175m axis"), 
           fill="#FFFFFFAA", label.size=0) +
  coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
              ylim = c(-1, 1)*clock_scale, 
              expand=TRUE) +
  scale_color_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
  theme_void() +
  theme(legend.position = "right", 
        legend.title = element_blank(),
        plot.background = element_rect(fill="white", color="white")) +
  guides(colour = guide_legend(override.aes = list(size=4, alpha=1), drop=FALSE)) +
  aes(x=x, y=y, color=resp_class, label=feature)

clockplot_complete_gp <- clockplot_base_gp +
  geom_point(aes(shape=error_type), alpha=0.5, size=2,
             data = clockplot_complete_df) +
  scale_shape_manual(breaks = c("Good MF found", "Bad MF included",
                                "Good MF missed", "Bad MF avoided"), 
                     values = c(19, 3, 4, 1))

ggsave(plot = clockplot_complete_gp, 
       filename = "clockplot_complete.png", 
       device = "png", path = "manuscript/figures",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

>>>>

A different story emerged, however, when the bad MFs were removed from this analysis. Good features were most commonly found to have their highest concentrations at the DCM or the surface, rather than being fixed with respect to depth. Of the 182 good MFs, less than a fifth had no trend with depth (44/249) and a majority had unequivocally lowest values in the 175 meter samples (those with 15m/DCM > 175m, 145 features). The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad.

Additionally, a large number of features manually identified as bad nonetheless had significant differences with depth. This was surprising because we had assumed that bad MFs corresponded to instrument or chemical noise, which we did not expect to have any biological trend. Further investigation of a few randomly selected bad features with a biological difference revealed the reason behind this: most of those investigated were actually tails of other MFs. Integrating just the tail of a peak retains the biological signal of the full peak while still looking visually like instrument noise, thereby introducing pseudoreplication in the feature space.

@@

```{r Fig5 clockplot}
resp_class_levels <- c("All equal", 
    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m")
clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
  mutate(resp_class=ifelse(resp_class=="Hard to tell", "All equal", resp_class)) %>%
  mutate(resp_class=factor(resp_class, levels=resp_class_levels))

@@

error_order <- c("Good MF found", "Good MF missed", "Bad MF included", 
                 "Bad MF avoided")
error_pchs <- c(19, 4, 3, 1)
error_unicodes <- c("\u25cf", "\u00d7", "\u002B", "\u25e6")
error_type_labels <- error_order %>%
  str_replace(" (?=[^ ]*$)", "\n") %>%
  paste0(" (", error_unicodes, ")")
# plot(1:4, pch=error_pchs, ylim=c(1, 6), xlim=c(-1.5, 4))
# points(1:4, 2:5, pch=error_unicodes)
# text(1:4, 3:6, label=error_order, adj = 1)

clockplot_hist_df <- clockplot_complete_df %>% 
  with(table(resp_class, error_type)) %>% 
  as.data.frame() %>%
  mutate(Freq=ifelse(Freq==1365, 60, Freq)) %>%
  group_by(error_type) %>%
  mutate(hjust_custom=ifelse(Freq>max(Freq)*2/3, 1.2, -0.2)) %>%
  mutate(text_color=ifelse(Freq>max(Freq)*2/3, "white", "black")) %>%
  ungroup() %>%
  mutate(freq_label=ifelse(Freq==0, "", Freq)) %>%
  mutate(freq_label=ifelse(
    resp_class=="All equal" & error_type=="Bad MF avoided", "1365 \u25ba", freq_label
  )) %>%
  mutate(hjust_custom=ifelse(
    resp_class=="All equal" & error_type=="Bad MF avoided", 1, hjust_custom
  )) %>%
  mutate(resp_class=factor(resp_class, levels=rev(levels(resp_class)))) %>%
  mutate(error_type=factor(error_type, levels=error_order,
                           labels=error_type_labels))
clockplot_hist_gp <- clockplot_hist_df %>% 
  ggplot() +
  geom_col(aes(y=resp_class, x=Freq, fill=resp_class)) +
  geom_text(aes(y=resp_class, x=Freq, hjust=hjust_custom, label=freq_label, color=text_color)) +
  scale_fill_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
  scale_color_identity() +
  scale_y_discrete(breaks=resp_class_levels) +
  facet_wrap(~error_type, nrow=1, scales="free_x") +
  theme_bw() +
  labs(x=NULL, y=NULL) +
  theme(legend.position = "none",
        axis.text.y = ggtext::element_markdown(color = c(
          "black", color_scale[2:13]
          ), face = "bold"))

clockplot_base_gp <- ggplot() +
  annotate("segment", x=c(0,0,0), y=c(0,0,0), 
           xend=axis_xends, yend=axis_yends,
           arrow=arrow(length=unit(0.02, "npc"), type = "closed")) +
  annotate("segment", x=c(0,0,0), y=c(0,0,0),
           xend=-axis_xends, yend=-axis_yends,
           lty = 2) +
  annotate("label", x=axis_xends, y=axis_yends+axis_y_dodge,
           label=c("15m axis", "DCM axis", "175m axis"), 
           fill="#FFFFFFAA", label.size=0, hjust=c(0.5, 0.9, 0.5)) +
  coord_fixed(xlim = c(-0.65, 1.05)*clock_scale, 
              ylim = c(-1, 1)*clock_scale, 
              expand=TRUE) +
  scale_color_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
  theme_void() +
  theme(legend.title = element_blank(),
        legend.position = c(0.75, 0.12),
        # legend.position="left",
        # legend.background = element_rect(fill="transparent", color="grey40"),
        plot.background = element_rect(fill="white", color="transparent"),
        plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "in")) +
  guides(colour = "none") +
  aes(x=x, y=y, color=resp_class, label=feature) +
  geom_point(aes(shape=error_type), alpha=0.5, size=2, data = clockplot_complete_df) +
  scale_shape_manual(breaks = error_order, values = error_pchs)

norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
#Palythine = FT1946
#Missed good MF w DCM minimum = FT1564 or FT1584
#DMS-Ac = FT0567
#Arsenobetaine = FT1432
#Interesting 146.1175 mass = FT1051

box_list <- list(c("FT1946", "FT1584"), c("FT0567", "FT1051")) %>%
  lapply(function(feat_ids){
    norm_area_data %>%
      filter(feature%in%feat_ids) %>%
      left_join(clockplot_complete_df %>% distinct(feature, resp_class),
                by="feature") %>%
      mutate(resp_class=factor(resp_class, levels=resp_class_levels)) %>%
      arrange(resp_class) %>%
      left_join(file_data %>% select(filename, depth), by="filename") %>%
      mutate(depth=factor(depth, levels=rev(levels(depth)))) %>%
      mutate(feature=case_when(
        feature=="FT1946"~"Palythine",
        feature=="FT1432"~"Arsenobetaine",
        feature=="FT0567"~"DMS-Ac",
        TRUE~feature
      )) %>%
      mutate(feature=factor(feature, levels=unique(feature))) %>%
      ggplot() +
      geom_boxplot(aes(x=norm_area, y=depth, color=resp_class)) +
      facet_wrap(~feature, ncol=1, scales = "free_x") +
      scale_x_log10(labels=scales::label_log(), breaks=10^(4:10)) +
      scale_color_manual(values = color_scale, breaks = names(color_scale)) +
      theme_bw() +
      theme(legend.position = "none",
            plot.margin=margin(l = 0, unit = "in")) +
      labs(x="Peak area", y=NULL)
})

clockplot_withboxplots <- gridExtra::arrangeGrob(
  box_list[[1]], clockplot_base_gp, box_list[[2]], 
  layout_matrix = matrix(c(1,2,2,2,3),nrow = 1)
)
clockplot_complete_gp <- gridExtra::arrangeGrob(
  clockplot_hist_gp, clockplot_withboxplots, 
  layout_matrix = matrix(c(1,2,2), ncol=1)
)
ggsave(plot = clockplot_complete_gp, 
       filename = "clockplot_complete.png", 
       device = "png", path = "manuscript/figures",
       width = 6.5, height = 7, units = "in", dpi = 300)
```
